---
title: "statsreasoning6"
format: pdf
pdf: default
gmf: default
editor: visual
---

## Piper Miller 

------------------------------------------------------------------------

# 1. Generalized linear models

```{r}
library(tidyverse) # For data wrangling
library(brms) # For stats
library(ggeffects) # for plotting model predictions
# Note: I needed to also install the `insight` and `see` packages to get `modelbased` to install properly; try that if you get similar error messages 
# install.packages('modelbased') # if you need to install this package
library(modelbased) # for plotting model predictions. supports the link scale (ggeffects does not)
# install.packages('faraway') # if you need to install this package
library(faraway) # For data on galapagos species richess
```



------------------------------------------------------------------------

### Conceptual practice

------------------------------------------------------------------------

#### Q1.1a What values can each of the response variables take on?

1.  Non negative integers
2.  0/1
3.  Real numbers between 0 and 100
4.  Real numbers
5.  Non negative real numbers

#### Q1.1b Choose a distribution that fits each of the response variables

1.  Poisson
2.  Bernoulli
3.  Beta
4.  Gaussian
5.  Gamma

------------------------------------------------------------------------

#### Q1.2 Choose a distribution that fits your final project response variable

Response variable: Abundance (counts) of harbor seals or stellar sea lions at Ano Nuevo State Park

Values: Non negative integers (whole numbers)

Distribution: Poission

1.2 GLM with a log link

### Explore the data

```{r}
# Read in the pre-stored data
data("gala")
# Check out the first 6 rows
head(gala)
# Look at the help page too!
```

------------------------------------------------------------------------

#### Q1.3 Plot a histogram of the response variable `Endemics`

Plot a histogram or density plot of the response variable `Endemics` to see how it is distributed. Does it look like a normal (Gaussian) distribution, which is a nice, symmetrical distribution that can take any value, or does it have certain constraints? What are those constraints?

```{r}
ggplot(gala,aes(x=Endemics)) +
  geom_histogram(binwidth = 1) +
  labs(x="Endemics",
       y="Frequency")
```

The distribution is skewed to the left. The response variable is constrained to non negative integers.

------------------------------------------------------------------------

#### Q1.4 Plot Endemics \~ Elevation

```{r}
ggplot(gala, aes(x=Elevation, y=Endemics)) +
  geom_point() +
  geom_smooth()
  labs(x="Elevation",
       y="Endemics")
```

------------------------------------------------------------------------

### Run the model

```{r}
# Endemics ~ Elevation
m.elev <- 
  brm(data = gala, # Give the model the penguins data
      # Choose a poisson distribution - THIS IS THE NEW PART!
      family = poisson(link = "log"),
      # Specify the model here. 
      Endemics ~ 1 + Elevation,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "output/m.elev")
```

And let's check out the summary:

```{r}
plot(m.elev)
pairs(m.elev)
summary(m.elev)
```

------------------------------------------------------------------------

#### Q1.5 Evaluate the output

The model did not run great, rhat is very high with a value of 2.96. The chains do not converge and overlap each other, the posterior distributions are very skewed and has extreme outliers. The CI values are not non zero and overlap with zero.

------------------------------------------------------------------------

#### Q1.6 Center the predictors

To fix convergence issues, it helps to center the predictors, increase the iterations and warmups, and set some priors. Please 1) center Elevation by creating a new column, `Elevation_ctr`, that subtracts each value in the `Elevation` column by the mean value (hint: `Elevation - mean(Elevation)`), 2) change the brm chain arguments to iter=6000, warmup=4000, 3) add some priors that seem reasonable, and 4) re-fit the model. Please call the new model m.elev2, then use plot(), summary(), and pairs() on it. This is a lot on your own, but you can do it! And we're here to help.

Remember, the model output rounds to two decimal places, so if the values are smaller than two decimal places, you can replace `summary(model)` with `print(model, digits = 4)`.

```{r include = FALSE}
gala <- gala |>
  mutate(Elevation_ctr = Elevation - mean(Elevation))

# Endemics ~ Elevation
m.elev2 <- 
  brm(data = gala, # Give the model the penguins data
      # Choose a poisson distribution - THIS IS THE NEW PART!
      family = poisson(link = "log"),
      # Specify the model here. 
      Endemics ~ 1 + Elevation_ctr,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 6000, warmup = 4000, chains = 4, cores = 4,
      prior = prior(normal(0, 0.1), class = b),
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "output/m.elev2")

```

### Plot the posterior

Before we get into interpreting the GLM output, let's plot the posterior prediction on the measurement scale using the `estimate_expectation()` function from the `modelbased` package.

```{r}
preds <- estimate_expectation(m.elev2, by = 'Elevation_ctr')
plot(preds, show_data = TRUE) #posterior plotted, had to install the "see" package 
```

Remember how we have a log link? What that means is that the model ran a linear model on the **logged** values of our response, `Endemics`. So the output that we get from the model is reporting the values of our slope, intercept, etc on the log scale.

The `predict_response()` function by default back-transforms from the log scale (log(`Endemics`)) of the linear model to the measurement scale of `Endemics`; that's what you see above. To see the linear model on the log scale, however, we can set predict = 'link':

```{r}
predslog <- estimate_expectation(m.elev2, by = 'Elevation_ctr', predict = 'link')
plot(predslog)
```

------------------------------------------------------------------------

### Interpreting link scale coefficients

#### Q1.7 What is the percent change on the response scale?

1.  Number of Clarkias blooming as a function of temperature in Celsius: 1.09

For every 1 C degree increase in temperature, Clarkias blooming increases by 197.4%.

```{r}
exp(1.09)
(2.974274-1.00)*100
```

2.  Density of sea urchins per square meter in a quadrat as a function of number of sea otters: -2.5

For every additional otter, the density of sea urchins per square meter decreases by 91.8%.

```{r}
exp(-2.5)
(0.082085-1.00)*100
```

3.  Number of tomatoes per plant as a function of kg of fertilizer: 6.24

For every kg of fertilizer, there is an 51,185.9% increase of tomatoes per plant.

```{r}
exp(6.24)
(512.8585-1.00)*100
```

------------------------------------------------------------------------

------------------------------------------------------------------------

### DIY: Run a model of non-endemic species \~ distance from Santa Cruz Island

#### Q1.8 Create a non-endemic column

```{r}
gala$non_Endemics <- gala$Species - gala$Endemics #$ pulls out column names from the data in front of it, species minus edemics will give us the non endemic species 
```

```{r}
ggplot(gala, aes(x=Scruz, y=non_Endemics)) +
  geom_point()
```

------------------------------------------------------------------------

#### Q1.9 Run a model of non Endemics \~ distance from Santa Cruz Island

```{r}
# Endemics \~ distance from Santa Cruz Island 
m.edem.dist <- 
  brm(data = gala, # Give the model the gala data
      family = poisson(link = "log"), #log function
      non_Endemics ~ 1 + Scruz,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      file = "output/m.edem.dist")

```

------------------------------------------------------------------------

#### Q1.10 Evaluate the output

```{r}
summary(m.edem.dist)
plot(m.edem.dist)
```

The model looks like it ran correctly, the chains overlap each other and the posterior distrubtion has one clean peak. Rhat is 1 and the intercept does not overlap zero.

------------------------------------------------------------------------

#### Q1.11 Interpret the output

1.  What is the effect of distance from Santa Cruz Island on number of non-endemic species? Report the a) original output on the log scale, b) your backtransformed value, and c) the percent change that this translates to. Describe the effect using the proper units.

The original log scale slope is -0.01, when backtransformed, the value becomes 0.9900498. Subtracting 1.00 from this value and multiplying by 100 gives us a percentage of -0.99% \~ -1%. This means that for every 1 km increase away from Santa Cruz Island, there is an 1% decrease in non endemic species.

2.  Does it seem like the slope estimate is different from zero? Why?

It does not, the 95% CI values are -0.01 and 0.00. We cannot confidently say that the slope is different than zero since these CI values overlap 0.

------------------------------------------------------------------------

#### Q1.12 Plot the posterior

```{r}
preds <- estimate_expectation(m.edem.dist, by = 'Scruz')
plot(preds, show_data = TRUE) 
#response scale 
```

```{r}
predslog <- estimate_expectation(m.edem.dist, by = 'Scruz', predict = 'link')
plot(predslog)
#log link scale 
```

------------------------------------------------------------------------

## 1.3 GLM with a logit link

### Bring in turtle data

In some species, incubation temperature of eggs affects the sex ratio of males vs females at birth. This dataset, from the `faraway` package, has the number of male and female turtles hatched at different temperatures. Let's calculate the total hatched and the proportion female:

```{r}
turtle <- faraway::turtle %>% 
  mutate(total_turtles = male + female,
         proportion_female = female/total_turtles)
```

And let's plot it!

```{r}
turtle %>% 
  ggplot(aes(x = temp, y = proportion_female)) +
  geom_point()
```

------------------------------------------------------------------------

### Run model

Since these are proportion data, a binomial link is the right choice of distribution to use. Let's fit a model. In this case, we need to specify our integer value of "successes" that we are interested in modeling while also including the number of "trials". In this case, "trials" are the total number of turtles while "successes" are the total number of females.

```{r}
m.turt <- 
  brm(data = turtle, # Give the model the data
      # Choose a binomial distribution - THIS IS THE NEW PART!
      family = binomial(link = "logit"),
      # Specify the model here. 
      female | trials(total_turtles) ~ 1 + temp,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 4000, warmup = 1000, chains = 4, cores = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "output/m.turt")

summary(m.turt)
plot(m.turt)
#looks like it ran correctly! 
```

------------------------------------------------------------------------

### Interpret model and plot predictions

According to this model, temperature reduces the probability that a turtle will be female by 2.27 on the logit scale per degree Celsius. This is tricky to get into the response scale, so we won't get into that now! Here's some slightly janky code that shows, based on this model, the predicted number of females as a function of temperature. It's standardized at 10 trials, aka 10 total turtles, so a value of 9 means 9 out of 10 turtles are female, and thus there's a 90% probability of being female.y

```{r}
pred <- predict_response(m.turt, condition = c(total_turtles = 10))

plot(pred)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

# 2. Multilevel models

------------------------------------------------------------------------

Much of the data in ecological and evolutionary biology is clustered in some way. This clustering can be spatial, with some replicate measurements being taken closer to one another than others, such as conducting 10 transects at one site and another 10 transects and a different site. It can temporal, where a bunch of measurements are closer together in time compared to others, such as all within the same season. A single individual, such as with large, well-studied marine mammals, or test subjects in a lab, may be sampled multiple times. There are many other examples!

If we're not careful, we may incorrectly assume that all measurements are independent of one another, where in reality, the measurements that are closer together in space, time, etc, are likely more closely related to one another than to other measurements. Using multi-level models (aka mixed effects models, hierarchical models) can fix this and improve our estimates of our parameters.

For some simple terminology: in multi-level models our predictor variables can be either *fixed effects* or *random effects*

-   Fixed effects are the variables we care about getting a parameter estimate from, like a slope value of how x affects y. **Every model we have run so far has used exclusively fixed effects**
-   Random effects are the variables that contribute to clustering in our data. They are many coefficients of the same type, and so we can help the model learn about them as a set.
-   Sometimes, a variable can be treated as either a random or a fixed effect.

------------------------------------------------------------------------

### Conceptual practice

#### Q2.1 Fixed effects vs random effects

1.  Student high school graduation rates as a function of: parental income, state of residence, and school district

    Fixed effects: Parental income

    Random effects: State of residence, school district

2.  Density of kelp as a function of: latitude, site, transect number, and density of sea urchins

    Fixed effects: Density of sea urchins

    Random effects: Latitude, site, transect number

3.  Probability of whale giving birth as a function of: age, annual temperature, year, individual ID

    Fixed effects: annual temperature, age

    Random effects: year, individual ID

------------------------------------------------------------------------

### Incorporate random effects

Let's go back to fiddler crabs and revisit the relationship between `air_temp` and crab `size`:

```{r}
pie_crab <- lterdatasampler::pie_crab %>% 
  mutate(site = as.factor(site))
```

```{r}
pie_crab %>% 
  ggplot(aes(x = air_temp, y = size)) +
  geom_point() #relationship between air temp and crab size 
```

------------------------------------------------------------------------

```{r}
m.watertemp <- 
  brm(data = pie_crab, # Give the model the crab data
      # Use a gamma distribution
      family = Gamma(link = "log"),
      # Specify the model here. 
      size ~ 1 + water_temp,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "output/m.watertemp")

print(m.watertemp, digits = 3) #ran well
```

Now let's incorporate site as a random effect:

```{r}
m.watertemp.site <- 
  brm(data = pie_crab, # Give the model the penguins data
      # Use a gamma distribution
      family = Gamma(link = "log"),
      # Specify the model here. 
      size ~ 1 + water_temp + (1|site),
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "output/m.watertemp.site")

print(m.watertemp.site, digits = 3) #ran well 
```

------------------------------------------------------------------------

#### Q2.2 What is the effect of water_temp on crab size on the response scale?

1.  The original output on the log scale is -0.039 when looking at the effect of water temp on crab size. The backtransformed output is 0.96 and the percentage change is -3.82%. This means that for every 1C increase in water temp, there an 3.82% decrease in crab size.

    ```{r}
    exp(-0.039)
    (0.9617507-1.00)*100
    ```

2.  It does mean like the slope estimate is different from zero. The 95% upper and lower CI values do not overlap zero meaning that we can confidently say that the slope estimate is different than zero.

------------------------------------------------------------------------

#### Q2.3 Compare WAIC and PSIS of the two models

The model including site has better predictive power, it has lower WAIC and PSIS values of the two models.

```{r}
loo(m.watertemp.site)
waic(m.watertemp.site)

loo(m.watertemp)
waic(m.watertemp)
```

------------------------------------------------------------------------

### Predict response of random effects

From the lecture, here's some code about how to show variation around each level of each random effect:

```{r}
preds <- predict_response(m.watertemp.site,
                          interval = "prediction",
                          terms = "site",
                          type = "random")

plot(preds)

```

------------------------------------------------------------------------

------------------------------------------------------------------------

### 
